---
title: "Workflow of pharmacokinetic modeling (model construction + data manipulation + visualization)"
author: "Nan-Hung Hsieh"
date: "2019/12/09 (update: `r Sys.Date()`)"
output: 
  html_document:
  fig_caption: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('..')
wd <- getwd()
knitr::opts_knit$set(root.dir =  wd)
```

```{R message=FALSE, warning=FALSE}
library(simuloR)
library(ggplot2)
library(pksensi)
library(httk)
theme_set(theme_bw())
```

```{r, eval=F}
View(Theoph)
```


```{r}
Theoph.MW <- 180.164 #g/mol
Theoph$conc.uM<-Theoph$conc * 1000 / Theoph.MW #uM
```


```{r}
plot(Theoph$Time, Theoph$conc.uM, col = Theoph$Subject)
```

```{r}
ggplot(Theoph, aes(x = Time, y =conc.uM, color = Subject)) + 
  geom_point() + geom_line() + labs(x = "Time (hr)", y = "Concentration (mg/L)")
```

```{r}
unique(Theoph$Dose)
```

```{r}
params <- parameterize_1comp(chem.name='Theophylline', species='Human')
out_httk <- solve_1comp(parameters=params, daily.dose = 5)
```

```{r}
out_df<- as.data.frame(out_httk)
plot(Theoph$Time, Theoph$conc.uM, col = Theoph$Subject)
lines(out_df$time*24, out_df$Ccompartment)
```

# Uncertainty analysis

```{r}
ffpk.mtc.out <- mcsim("models/ffpk.model", "inputs/ffpk.mtc.in")
```


```{r}
head(ffpk.mtc.out)
```

```{r}
par(mfrow = c(1,4))
hist(ffpk.mtc.out$Fgutabs, main = "Fgutabs", xlab = "", ylab = "")
hist(ffpk.mtc.out$kgutabs, main = "kgutabs", xlab = "", ylab = "")
hist(ffpk.mtc.out$kelim, main = "kelim", xlab = "", ylab = "")
hist(ffpk.mtc.out$Vdist, main = "Vdist", xlab = "", ylab = "")
```

```{r}
# Find location of the first and last time points by which()
index <- which(names(ffpk.mtc.out)=="Ccompartment_1.1" | names(ffpk.mtc.out)=="Ccompartment_1.11")
# Use apply() to find the quantile of each time point
X <- apply(ffpk.mtc.out[,index[1]:index[2]], 2, quantile, c(0.5, 0.01,0.99))
dat <- t(X)
# Set column names
colnames(dat) <- c("median", "LCL", "UCL")
df <- as.data.frame(dat)
# Set corresponding time
df$time <- c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 10, 12, 24)
df
```

```{r}
t <- c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 10, 12, 24)
str.pt <- which(names(ffpk.mtc.out)=="Ccompartment_1.1") 
end.pt <- which(names(ffpk.mtc.out)=="Ccompartment_1.11") 
par(mfrow = c(1,1))
plot(Theoph$Time, Theoph$conc.uM, pch = "")
for(i in 1:1000){
  lines(t, ffpk.mtc.out[i,str.pt:end.pt], col="grey")
}
points(Theoph$Time, Theoph$conc.uM, col = Theoph$Subject)
```


# pksensi

1. Set parameter distributions (assign parms, dist, q.qarg)
```{r}
parameters <- c("Fgutabs",  "kgutabs", "kelim", "Vdist")  
dist <- c("Uniform", "LogNormal", "LogNormal", "LogNormal") # MCSim definition
q.arg <- list(list(0.8, 1.0),
            list(2.18, 3),
            list(0.02183918, 2),
            list(5.012049, 3))
```

2. Set experiment time-points, output variables, and conditions

```{r}
outputs <- c("Ccompartment")
times <- c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 10, 12, 24)
conditions <- c("MW = 180.164", "Period = 48", "IngDose = 5.0")
```

3. Modeling

```{r}
mcmc_out <- makemcsim("models/pbtk1cpt.model")
```

```{r}
set.seed(2222)
y<-solve_mcsim(mName = "models/pbtk1cpt.model", params = parameters, vars = outputs, monte_carlo = 1000,
               dist = dist, q.arg = q.arg, time = times, condition = conditions)
```

```{r}
pksim(y)
points(Theoph$Time, Theoph$conc.uM, col = Theoph$Subject)
```



# Calibration

```{r}
set.seed(1)
ffpk.mcmc.out <- mcsim(model = "models/ffpk.model", input = "inputs/ffpk.mcmc.in")
```


```{r}
par(mfrow = c(2,2))
plot(ffpk.mcmc.out$Fgutabs.1., type = "l")
plot(ffpk.mcmc.out$kgutabs.1., type = "l")
plot(ffpk.mcmc.out$kelim.1., type = "l")
plot(ffpk.mcmc.out$Vdist.1., type = "l")
```



```{r}
check.out <- read.delim("MCMC.check.out") 
plot(check.out$Data, check.out$Prediction, xlim=c(0,60), ylim=c(0,60))
abline(0,1)
```


Use R
```{r}
IngDose <- 4.02
Fgutabs <- ffpk.mcmc.out$Fgutabs.1.
kgutabs <- ffpk.mcmc.out$kgutabs.1.
kelim <- ffpk.mcmc.out$kelim.1.	
Vdist <- ffpk.mcmc.out$Vdist.1.
MW <- 180.164

t <- Theoph$Time[1:11]
obs<- Theoph$conc.uM[1:11]
A <- (IngDose * Fgutabs * kgutabs)/(Vdist * (kgutabs - kelim))

Conc.matrix <- matrix(nrow = length(Fgutabs), ncol = length(t))
for(i in 1:11){
  Conc.matrix[,i] <- A * exp(-kelim * t[i]) - A * exp(-kgutabs * t[i])
}
Ccompartment <- Conc.matrix / MW * 1000
plot(t, obs, pch = "")
for(i in 501:1000){
  lines(t, Ccompartment[i,], col="grey")
}
points(t, obs)
```


```{r}
plot(obs, obs, pch = "")
for(i in 501:1000){
  points(obs, Ccompartment[i,], pch = "+")
}
abline(0,1)
```

```{r}
plot(obs, obs, ylim = c(0.1, 10), pch = "", log="y")
for(i in 501:1000){
  points(obs, obs/Ccompartment[i,], pch = "+")
}
abline(0,0)
abline(log(2,10), 0, lty=2)
abline(log(0.5,10), 0, lty=2)
```

Use setpoints
```{r}
ffpk.setpts.out <- mcsim("models/ffpk.model", input = "inputs/ffpk.setpts.in")
```

```{r}
t <- Theoph$Time[1:11]
obs<- Theoph$conc.uM[1:11]
str.pt <- which(names(ffpk.setpts.out)=="Ccompartment_1.1") 
end.pt <- which(names(ffpk.setpts.out)=="Ccompartment_1.11") 
par(mfrow = c(1,1))
plot(t, obs, pch = "")
for(i in 501:1000){
  lines(t, ffpk.setpts.out[i,str.pt:end.pt], col="grey")
}
points(t, obs)
```

Use setpoints - Global evaluation



Compare the CSS

```{r}
library(httk)
calc_analytic_css(chem.name = "theophylline", model = "1compartment")
```


```{r}
parameters <- parameterize_1comp(chem.name = "theophylline")
parameters$Fgutabs <- median(ffpk.mcmc.out$Fgutabs.1.) / parameters$hepatic.bioavailability
parameters$kgutabs <- median(ffpk.mcmc.out$kgutabs.1.)
parameters$kelim <- median(ffpk.mcmc.out$kelim.1.)
parameters$Vdist <- median(ffpk.mcmc.out$Vdist.1.)
unlist(parameters)
calc_analytic_css(parameters = parameters, model = "1compartment")
```



```{r}
sessionInfo()
```

# Reference

Bois F.Y. 2013. [Bayesian Inference](https://link.springer.com/protocol/10.1007/978-1-62703-059-5_25#Bib1). In: Reisfeld B., Mayeno A. (eds) Computational Toxicology. Methods in Molecular Biology (Methods and Protocols), vol 930. Humana Press, Totowa, NJ
