---
title: "Exercise 1.1: Exploratory data analysis"
author: "Nan-Hung Hsieh"
date: "2019/12/09 (update: `r Sys.Date()`)"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
  fig_caption: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = T)
```


# Task 

Learn how to use R to conduct basic analysis for PK data. Now, we have a theophylline PK dataset. The purpose of this exercise is to develop the simple PK model and use it to describe the PK of theophylline. First, look into the theophylline dataset. The Theoph data frame has 132 rows and 5 columns of data from an experiment on the pharmacokinetics of theophylline.

Questions  
- Where can I get help?  
- How do I read data into R?  
- What is a data frame?  
- How do I assign variables?  
- How do I access subsets of a data frame?  
- How do I calculate simple statistics like mean and median?  
- How can I plot my data?  

Objectives  
- Read tabular data from a file into a program.  
- Assign values to variables.  
- Select individual values and subsections from data.  
- Perform operations on a data frame of data.  
- Display simple graphs (base and ggplot).  
- Find the Cmax and Tmax for each individual.  


# Solution

## Basic data analysis

use `getwd()` to find where you are.
```{r}
getwd()
```

Use `?` before function to get help and look into the `Theoph` dataset.
```{r}
?Theoph
```

Use `head()` function to check the top row of theophylline PK data.
```{r}
head(x = Theoph)
```

Use `tail()` function to check the top row of theophylline PK data.
```{r}
tail(x = Theoph)
```

Use `class()` to see the data type of Theophylline
```{r}
class(x = Theoph)
```

Use `View()` function to overview the data 
```{r. eval=F}
View(x = Theoph)
```

Use  `write.csv()` to save your data to a csv file and named it "Theoph.csv"
```{r}
write.csv(x = Theoph, file = "Theoph.csv")
```


Now, there is a csv file called "Theoph.csv". Use `read.csv()` to load it to your wrok space. In addtion, assign the data and named it `dat`

```{r}
dat <- read.csv(file = "Theoph.csv")
```

Sometimes you might need to focus on specific subgroup (individual). Use `subset()` to select Subject 1.
```{r}
subset(Theoph, Subject == 1)
```

Assign the variable `S1` to store the information of Subject 1. Then, try using `$` and `[]` to select the blood concentraion
```{r}
S1 <- subset(x = Theoph, Subject==1)
S1$conc
```

How's the average concentration of Subject 1?
```{r}
mean(S1$conc)
```

How's the maximum concentration ($Cmax$) of Subject 1?
```{r}
max(S1$conc)
```

Plot the PK profile for subject 1.
```{r}
plot(x = S1$Time, y = S1$conc)
```

Plot the PK profile for all subjects.
```{r}
plot(x = Theoph$Time, y = Theoph$conc, col = Theoph$Subject)
```

Try to seperate each individaul to different sub figures.
```{r}
par(mfrow=c(m=3,n=4))
for(i in 1:12){
  Si <- subset(x = Theoph, Subject == i)  # Select subject
  plot(x = Si$Time, y = Si$conc, main = i)   # plot PK profile
}
```


## Tidyverse package

Use `library()` to load `tidyverse`
```{r}
library(tidyverse)
```

Use `ggplot` to generate the PK profile for all subjects.
```{r}
Theoph %>% ggplot(aes(x = Time, y =conc, color = Subject)) + 
  geom_point()
```

Look into the data tye of `Theoph$Subject`
```{r}
class(Theoph$Subject)
```

The currnt order of the subject is based on the peak concentration. Try to reorder it by number.
```{r}
Theoph$Subject
```

Re-create the ggplot and check the different with previous plot.
```{r}
Theoph$Subject <- factor(Theoph$Subject, 
                            level = c("1", "2", "3", "4", "5", "6", 
                                      "7", "8", "9", "10", "11","12"))
```

Try to seperate each individaul to different sub figures.
```{r}
Theoph$Subject
```

Find the Tmax
- First, assign the peak concentration and named it `S1.Cmax`
```{r}
Theoph %>% ggplot(aes(x = Time, y =conc, color = Subject)) + 
  geom_point() + geom_line() + 
  labs(x = "Time (hr)", y = "Concentration (mg/L)")
```

- Second, use `which()` to find the corresponding row number and assigned the variable `Cmax.pt`
```{r}
Theoph %>% ggplot(aes(x = Time, y =conc)) + 
  geom_point() + geom_line() + 
  facet_wrap(~Subject) + 
  labs(x = "Time (hr)", y = "Concentration (mg/L)")
```

- Third, use `Cmax.pt` to find the corresponding $T_{max}$
```{r}
S1.Cmax <- max(S1$conc)
S1.Cmax
```


Use pipline `%>%`, `group_by()`, and `summarise()` to manipulate the data and find peak concentration for all subjects. Assigned the new variable `Cmax_pop`
```{r}
which(S1$conc == S1.Cmax)
```

Plot histogram and baplot
```{r}
Cmax.pt <- which(S1$conc == S1.Cmax)
T_max <- S1$Time[Cmax.pt]
T_max
```

Try to figure ou the relationship between body weight and peak concentration.
- Summarize the average body weight for each individual
```{r}
Cmax_pop <- Theoph %>% group_by(Subject) %>% summarise(C_max = max(conc))
Cmax_pop
```

```{r}
summary(Cmax_pop$C_max)
```

```{r}
hist(Cmax_pop$C_max) # histogram
```

```{r}
boxplot(Cmax_pop$C_max) # boxplot
```

- Visualize by ggplot
```{r}
Theoph %>% group_by(Subject) %>% 
  summarise(C_max = max(conc), Wt=mean(Wt)) %>% 
  ggplot(aes(x=Wt, y=C_max)) + geom_point()
```

- Add linear regression model
```{r}
Theoph %>% group_by(Subject) %>% 
  summarise(C_max = max(conc), Wt=mean(Wt)) %>% 
  ggplot(aes(x=Wt, y=C_max)) + geom_point() + geom_smooth(method = "lm")
```

Create the table that summarrize the $C_{max}$ and $T_{max}$ for each individual.
- First, generate data frame `df` that include subject number, $C_{max}$ and $T_{max}$ 
```{r}
df <- Theoph %>% group_by(Subject) %>% summarize(C_max = max(conc))
df$T_max <- NA
df
```

- Second, add find the corresponding $T_{max}$ for subject 1
```{r}
dat <- subset(Theoph, Subject==1)
C_max.pt <- which(dat$conc == df$C_max[1])
df$T_max[1] <- dat$Time[C_max.pt]
df
```

- Third, use for loop add find the corresponding $T_{max}$ for other subjects
```{r}
for(i in 2:12){
  dat <- subset(Theoph, Subject==i)
  C_max.pt <- which(dat$conc == df$C_max[i])
  df$T_max[i] <- dat$Time[C_max.pt]
}
df
```

