---
title: "Exercise 1.3 Parameter setting and simulation"
author: "Nan-Hung Hsieh"
date: "2019/12/09 (update: `r Sys.Date()`)"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
  fig_caption: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
setwd('..')
wd <- getwd()
knitr::opts_knit$set(root.dir =  wd)
```

# Task

- Extract the chemical information from httk package

- Use the parameters in the developed model and conduct the simulation 

- Compare the difference between data and the model simulation result (Cmax, Tmax)

# Solution

## Packages

```{R message=FALSE, warning=FALSE}
library(httk)
library(tidyverse)
library(simuloR)
```

## Simulation by R

```{r}
parms <- parameterize_1comp(chem.name = "theophylline")
parms
```

```{r}
ffpk <- function(IngDose, Fgutabs, kgutabs, kelim, Vdist, BW, t){
    A <- (IngDose * Fgutabs * kgutabs)/(Vdist * BW * (kgutabs - kelim))
    Conc <- A * exp(-kelim * t) - A * exp(-kgutabs * t)
    return(Conc)
}
```

```{r}
Fgutabs <- parms$Fgutabs * parms$hepatic.bioavailability
kgutabs <- parms$kgutabs
kelim <- parms$kelim
Vdist <- parms$Vdist
```


```{r}
S1 <- subset(Theoph, Subject == 1)
mean(S1$Dose)
mean(S1$Wt)
```

```{r}
IngDose <- mean(S1$Dose)*mean(S1$Wt)
IngDose
```


```{r}
out1 <- ffpk(IngDose = IngDose, 
             Fgutabs = Fgutabs,
             kgutabs = kgutabs, kelim = kelim,
             Vdist = Vdist, 
             BW = 79.6,
             t = S1$Time)
out1
```


```{r}
conc1 <- subset(Theoph, Subject == 1) %>% select(conc)

plot(S1$Time, conc1$conc)
lines(S1$Time, out1)
```

## Simulation by MCSim

### Non-Compartment model

```r
IngDose    = 319.992;     # ingested dose (mg)
BW         = 79.6;        # body weight (kg)
Fgutabs    = 0.9142;      # bioavailability (-) 
kgutabs    = 2.18;        # absortion rate (/h)
kelim      = 0.2654;      # elimination rate constant (/h)
Vdist      = 0.7802;      # volumes (L/kg BW)

Simulation {
  Print (Conc 0.00   0.25   0.57   1.12   2.02   3.82   5.10   7.03   9.05  12.12  24.37);
}
End.
```

**Use linux command**

- Create exectuable program (`makemcsim`)
```bash
makemcsim models/ffpk.model mcsim.ffpk
```
- Conduct simulation
```bash
models/mcsim.ffpk inputs/ffpk.in
```

**Use simuloR function**

- Create exectuable program 
```{r}
ffpk.model <- "models/ffpk.model"
makemcsim(model = ffpk.model)
```

- Create exectuable program and conduct simulation
```{r}
mcsim(model = "models/ffpk.model", input = "inputs/ffpk.in")
```

- Load file
```{r}
read.delim(file = "sim.out")
```

```{r}
read.delim(file = "sim.out", skip = 1)
```

### Compartment model

```
IngDose    = 319.992;     # ingested dose (mg)
BW         = 79.6;        # body weight (kg)
Period     = 48;          # period of the exposure/no exposure cycle (h)
Fgutabs    = 0.9142;      # bioavailability (-) 
kgutabs    = 2.18;        # absortion rate (/h)
kelim      = 0.2654;      # elimination rate constant (/h)
Vdist      = 0.7802;      # volumes (L/kg BW)

Simulation {
  Print (Ccpt 0.00   0.25   0.57   1.12   2.02   3.82   5.10   7.03   9.05  12.12  24.37);
}
End.
```

```{r}
out1 <- mcsim(model = "models/ffpk.model", input = "inputs/ffpk.in")
out1
```

```{r}
plot(x = S1$Time, y = conc1$conc)
lines(x = out1$Time, y = out1$Conc)
```
