---
title: "Exercise 2.1: Uncertainty analysis on PK model"
author: "Nan-Hung Hsieh"
date: "2019/12/09 (update: `r Sys.Date()`)"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
  fig_caption: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
setwd('..')
wd <- getwd()
knitr::opts_knit$set(root.dir =  wd)
```

# TASK

- In the previous exercise, we find that the predcited result can not used to describe the real cases. Therefore, we need to conduct the uncertainty analysis to figure out how to reset the model parameter.

- Use the parameter distributions that we test in uncertainty analysis and conduct MCMC simulation to do model calibration.

# Basic MOnte Carlo practice

```{r}
x <- rnorm(n = 28, mean = 10, sd =  2)
x
```

```{r}
mean(x)
sd(x)
```

```{r}
hist(x)
```

**Generating random number**

```{r}
runif(n=5, min=0, max=1)
runif(n=5, min=0, max=1)
set.seed(1)
runif(n=5, min=0, max=1)
set.seed(1)
runif(n=5, min=0, max=1)
```


```{r}
x <- runif(10)
plot(x)
hist(x)
```

```{r}
x <- runif(100)
plot(x)
hist(x)
```

```{r}
x <- runif(1000)
plot(x)
hist(x)
```

```{r}
x <- runif(10000)
plot(x)
hist(x)
```

# Monte Carlo method in GNU MCSim

## [`MonteCarlo()` specification](https://www.gnu.org/software/mcsim/mcsim.html#MonteCarlo_0028_0029-specification)

MonteCarlo specification gives general information required for the runs: the output file name, the number of runs to perform, and a starting seed for the random number generator. Its syntax is:

```r
MonteCarlo("<OutputFilename>", <nRuns>, <RandomSeed>);
```

The character string <OutputFilename>, enclosed in double quotes, should be a valid filename for your operating system. If a null-string "" is given, the default name simmc.out will be used. The number of runs <nRuns> should be an integer, and is only limited by either your storage space for the output file or the largest (long) integer available on your machine. When using the GNU Scientific Library, the seed <RandomSeed> of the pseudo-random number generator can be any positive integer (including zero). When using the native code, it can be any positive real number (seeds between 1.0 and 2147483646.0 are used as is, the others are rescaled silently within those bounds). Here is an example of use:

```r
MonteCarlo("percsimmc.out", 50000, 9386);
```

## [`Distrib()` specification](https://www.gnu.org/software/mcsim/mcsim.html#Distrib_0028_0029-specification)

`Distrib()` specifies the name of the parameter to sample, and its sampling distribution. Its syntax is:

```r
Distrib(<parameter identifier>, <distribution-name>, [<shape parameters>]);
```

For example
`Normal`, takes two reals numbers as parameters: the mean and the standard deviation, the latter being strictly positive.
`LogNormal`, takes two reals numbers as parameters: the geometric mean (exponential of the mean in log-space) and the geometric standard deviation (exponential, strictly superior to 1, of the standard deviation in log-space).


# SOLUTION

## Packages

```{R message=FALSE, warning=FALSE}
library(simuloR)
library(tidyverse)
library(pksensi)
library(httk)
```

## Uncertainty analysis of PK model in R

### Model

```{r}
ffpk <- function(IngDose, Fgutabs, kgutabs, kelim, Vdist, BW, t){
    A <- (IngDose * Fgutabs * kgutabs)/(Vdist * BW * (kgutabs - kelim))
    Conc <- A * exp(-kelim * t) - A * exp(-kgutabs * t)
    return(Conc)
}
```

### Parameters

```{r}
parms <- httk::parameterize_1comp(chem.name = "theophylline")
```

```{r}
Fgutabs <- parms$Fgutabs * parms$hepatic.bioavailability
kgutabs <- parms$kgutabs
kelim <- parms$kelim
Vdist <- parms$Vdist
```

### Monte Carlo simulation

```{r}
LL <- 0.2
Fgutabs_LL <- 0.8
kgutabs_LL <- kgutabs * LL
kelim_LL <- kelim * LL
Vdist_LL <- Vdist * LL
c(Fgutabs_LL, kgutabs_LL, kelim_LL, Vdist_LL)
```


```{r}
UL <- 5
Fgutabs_UL <- 1
kgutabs_UL <- kgutabs * UL
kelim_UL <- kelim * UL
Vdist_UL <- Vdist * UL
c(Fgutabs_UL, kgutabs_UL, kelim_UL, Vdist_UL)
```


```{r}
n <- 1000
set.seed(1)
dist_Fgutabs <- runif(n, Fgutabs_LL, Fgutabs_UL)
set.seed(1)
dist_kgutabs <- runif(n, kgutabs_LL, kgutabs_UL)
set.seed(1)
dist_kelim <- runif(n, kelim_LL, kelim_UL)
set.seed(1)
dist_Vdist <- runif(n, Vdist_LL, Vdist_UL)
```


```{r}
df <- data.frame(dist_Fgutabs, dist_kgutabs, dist_kelim, dist_Vdist)
head(df)
```


```{r}
names(df) <- c("Fgutabs", "kgutabs", "kelim", "Vdist")
names(df)
```


```{r}
par(mfrow=c(1,4))
hist(df$Fgutabs)
hist(df$kgutabs)
hist(df$kelim)
hist(df$Vdist)
```

```{r}
S1 <- subset(Theoph, Subject==1)
t1 <- S1$Time
Dose1 <- mean(S1$Dose)
BW1 <- mean(S1$Wt)
```

```{r}
m.Conc <- matrix(nrow=n, ncol=11)
head(m.Conc)
```

```{r}
colnames(m.Conc) <- t1
head(m.Conc)
```

```{r}
dim(m.Conc)
```


```{r}
m.Conc[1,] <- ffpk(IngDose = Dose1*BW1, 
     Fgutabs = df$Fgutabs[1], kgutabs = df$kgutabs[1], 
     kelim = df$kelim[1], 
     Vdist =  df$Vdist[1],
     BW = BW1,
     t = t1)
head(m.Conc)
```


```{r}
for(i in 2:n){
  m.Conc[i,] <- ffpk(IngDose = Dose1*BW1, 
                     Fgutabs = df$Fgutabs[i], kgutabs = df$kgutabs[i], 
                     kelim = df$kelim[i], 
                     Vdist =  df$Vdist[i],
                     BW = BW1,
                     t = t1)
}
tail(m.Conc)
```


### Visualization

```{r}
plot(t1, m.Conc[1,], type="l")
```

```{r}
max.conc <- max(m.Conc)
plot(t1, m.Conc[1,], type="l", col="grey", ylim = c(0, max.conc))
for(i in 2:n){
  lines(t1, m.Conc[i,], col="grey")
}
```

```{r}
Conc1 <- S1$conc
plot(t1, Conc1)
max.conc <- max(m.Conc)
plot(t1, m.Conc[1,], type="l", col="grey", ylim = c(0, max.conc))
for(i in 2:n){
  lines(t1, m.Conc[i,], col="grey")
}
points(t1, Conc1)
```

## Uncertainty analysis of PK model in GNU MCSim

### simuloR

**Input file**

```r
MonteCarlo ("simmc.out", 
            1000, 
            10101010);

# Constant
IngDose    = 319.992;     # ingested dose (mg)
BW         = 79.6;        # body weight (kg)

# Distribution
Distrib (Fgutabs, Uniform, 0.8, 1);              # bioavailability (-) 
Distrib (kgutabs, LogUniform, 0.218, 21.8);      # absortion rate (/h)
Distrib (kelim, Uniform, 0.05307864, 1.326966);  # elimination rate constant (/h)
Distrib (Vdist, Uniform, 0.15603994, 3.900998);  # volumes (L/kg BW)

Simulation {
  Print (Conc 0.0   0.25   0.57   1.12   2.02   3.82   5.10   7.03   9.05  12.12  24.37);
}
End.
```

```{r}
set.seed(2222)
ffpk.mtc.out <- mcsim("models/ffpk.model", "inputs/ffpk.mtc.in")
```


```{r}
head(ffpk.mtc.out)
```

```{r}
par(mfrow = c(1,4))
hist(ffpk.mtc.out$Fgutabs, main = "Fgutabs", xlab = "", ylab = "")
hist(ffpk.mtc.out$kgutabs, main = "kgutabs", xlab = "", ylab = "")
hist(ffpk.mtc.out$kelim, main = "kelim", xlab = "", ylab = "")
hist(ffpk.mtc.out$Vdist, main = "Vdist", xlab = "", ylab = "")
```


```{r}
hist(log(ffpk.mtc.out$kgutabs), main = "kgutabs", xlab = "", ylab = "")
```


```{r}
Conc.str <- which(names(ffpk.mtc.out)=="Conc_1.1")
Conc.end <- which(names(ffpk.mtc.out)=="Conc_1.10")
df.Conc <- ffpk.mtc.out[,Conc.str:Conc.end]
head(df.Conc)
```

```{r}
max.Conc <- max(df.Conc)
plot(t1[-1], df.Conc[1,], type="l", col="grey", ylim=c(0,max.Conc))
for(i in 2:1000){
  lines(t1[-1], df.Conc[i,], col="grey")
}
points(t1, Conc1)
```

### pksensi

1. Set parameter distributions (assign parms, dist, q.qarg)
```{r}
parameters <- c("Fgutabs",  "kgutabs", "kelim", "Vdist")  
dist <- c("Uniform", "LogUniform", "Uniform", "Uniform") # MCSim definition
q.arg <- list(list(0.8, 1.0), 
              list(0.218, 21.8),
              list(0.05307864, 1.326966),
              list(0.15603994, 3.900998))
```

2. Set experiment time-points, output variables, and conditions

```{r}
outputs <- c("Ccpt")
times <- c(0.00, 0.25, 0.57, 1.12, 2.02, 3.82, 5.10, 7.03, 9.05, 12.12, 24.37)
conditions <- c("BW = 79.6", "Period = 48", "IngDose = 319.992")
```

3. Modeling

```{r}
model <- "models/pbtk1cpt.model"
```

```r
makemcsim(model)
```

```{r}
set.seed(2222)
y<-solve_mcsim(mName = model, params = parameters, vars = outputs, monte_carlo = 10000,
               dist = dist, q.arg = q.arg, time = times, condition = conditions)
```

```{r}
pksim(y)
points(t1, Conc1)
```



