---
title: "Exercise 2.2 Model calibration"
author: "Nan-Hung Hsieh"
date: "2019/12/09 (update: `r Sys.Date()`)"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  fig_caption: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('..')
wd <- getwd()
knitr::opts_knit$set(root.dir =  wd)
```

## Task

- After the uncertainty analysis, we can calibrate the model parameters by Markov chain Monte Carlo technique

- Use the parameter distributions that we test in uncertainty analysis and conduct MCMC simulation to do model

## Solution

```{R message=FALSE, warning=FALSE}
library(simuloR)
library(tidyverse)
library(httk)
library(corrplot)
library(bayesplot)
library(rstan)
theme_set(theme_bw())
color_scheme_set("mix-blue-red")
```


```r
# ffpk.mcmc.in ---------------------------------------------------------------

MCMC ("MCMC.default.out","", # name of output file
      "",         # name of data file
      2000,0,     # iterations, print predictions flag
      1,2000,     # printing frequency, iters to print
      10101010);  # random seed (default)
  
Level {
  
  Distrib (Fgutabs, Uniform, 0.8, 1);
  Distrib (kgutabs, LogUniform, 0.218, 21.8);
  Distrib (kelim, Uniform, 0.05307864, 1.326966); 
  Distrib (Vdist, Uniform, 0.15603994, 3.900998); 
  
  Likelihood(Conc, LogNormal, Prediction(Conc) , 1.1);
  
  Simulation { # 1
    IngDose  = 4.02;     # ingested dose (mg)
    Print (Conc, 0.01  0.25  0.57  1.12  2.02  3.82  5.10  7.03  9.05 12.12 24.37 ); 
    Data (Conc, 0.74  2.84  6.57 10.50  9.66  8.58  8.36  7.47  6.89  5.94  3.28 );
  }
}
End.
```

### Single chain simulation

```{r}
set.seed(1)
ffpk.mcmc.1 <- mcsim(model = "models/ffpk.model", input = "inputs/ffpk.mcmc.in")
```


```{r}
par(mfrow = c(2,2))
plot(ffpk.mcmc.1$Fgutabs.1., type = "l")
plot(ffpk.mcmc.1$kgutabs.1., type = "l")
plot(ffpk.mcmc.1$kelim.1., type = "l")
plot(ffpk.mcmc.1$Vdist.1., type = "l")
```


```{r}
parms <- httk::parameterize_1comp(chem.name = "theophylline")
Fgutabs <- parms$Fgutabs * parms$hepatic.bioavailability
kgutabs <- parms$kgutabs
kelim <- parms$kelim
Vdist <- parms$Vdist
```



```{r}
i <- 1001:2000
ffpk.mcmc.1$Fgutabs.1.[i] %>% density() %>% plot(main = "Fgutabs")
plot.rng <- seq(par("usr")[1], par("usr")[2], length.out = 1000)
prior.dens <- do.call("dunif", c(list(x=plot.rng), c(0.8,1)))
lines(plot.rng, prior.dens, lty="dashed")
abline(v=Fgutabs, col="red")
```


```{r}
ffpk.mcmc.1$kgutabs.1.[i] %>% density() %>% plot(main = "kgutabs")
plot.rng <- seq(par("usr")[1], par("usr")[2], length.out = 1000)
prior.dens <- do.call("dunif", c(list(x=plot.rng), c(0.218,21.8)))
lines(plot.rng, prior.dens, lty="dashed")
abline(v=kgutabs, col="red") # ka = 2.18 
```

```{r}
ffpk.mcmc.1$kelim.1.[i] %>% density() %>% plot(main = "kelim")
plot.rng <- seq(par("usr")[1], par("usr")[2], length.out = 1000)
prior.dens <- do.call("dunif", c(list(x=plot.rng), c(0.05307864, 1.326966)))
lines(plot.rng, prior.dens, lty="dashed")
abline(v=kelim, col="red") # ke = 0.2653932
```


```{r}
ffpk.mcmc.1$Vdist.1.[i] %>% density() %>% plot(main = "Vdist")
plot.rng <- seq(par("usr")[1], par("usr")[2], length.out = 1000)
prior.dens <- do.call("dunif", c(list(x=plot.rng), c(0.15603994, 3.900998)))
lines(plot.rng, prior.dens, lty="dashed")
abline(v=Vdist, col="red") # 0.78
```

```{r}
names(ffpk.mcmc.1)
ffpk.mcmc.1[i,c("Fgutabs.1.","kgutabs.1.","kelim.1.","Vdist.1.")] %>% cor()
```


```{r}
ffpk.mcmc.1[i,c("Fgutabs.1.","kgutabs.1.","kelim.1.","Vdist.1.")] %>% cor() %>% corrplot(method = "number")
```



```{r}
check.out <- read.delim("MCMC.check.out") 
plot(check.out$Data, check.out$Prediction, xlim=c(0,10), ylim=c(0,10))
abline(0,1)
```



```{r}
Observed <- check.out$Data
Expected <- check.out$Prediction
plot(Expected, Observed/Expected, pch='x')
abline(1,0)
```


```{r}
ffpk <- function(D, Fa, ka, ke, V, t){
    MW <- 180.167
    A <- (D * Fa * ka)/(V * (ka - ke))
    Conc <- A * exp(-ke * t) - A * exp(-ka * t)
    return(Conc)
}
```


```{r}
IngDose <- 4.02
Fgutabs <- ffpk.mcmc.1$Fgutabs.1.
kgutabs <- ffpk.mcmc.1$kgutabs.1.
kelim <- ffpk.mcmc.1$kelim.1.	
Vdist <- ffpk.mcmc.1$Vdist.1.

t <- Theoph$Time[1:11]
obs<- Theoph$con[1:11]


Conc.matrix <- matrix(nrow = length(Fgutabs), ncol = length(t))
for(i in 1:11){
  Conc.matrix[,i] <- ffpk(D = 4.02, Fa = Fgutabs,ka = kgutabs, ke = kelim, V = Vdist,t = t[i])
}
```


```{r}
tail(Conc.matrix)
```




```{r}
plot(t, obs, pch = "")
for(i in 501:1000){
  lines(t, Conc.matrix[i,], col="grey")
}
points(t, obs)
```



### Multi-chains simulation


```{r}
set.seed(2)
ffpk.mcmc.2 <- mcsim(model = "models/ffpk.model", input = "inputs/ffpk.mcmc.in")
set.seed(3)
ffpk.mcmc.3 <- mcsim(model = "models/ffpk.model", input = "inputs/ffpk.mcmc.in")
set.seed(4)
ffpk.mcmc.4 <- mcsim(model = "models/ffpk.model", input = "inputs/ffpk.mcmc.in")
```

Use mcmc_array() to create the 3-D array (iterations * chains * [parameters+4]) to store all outputs.

```{r}
sims <- mcmc_array(data = list(ffpk.mcmc.1, ffpk.mcmc.2, ffpk.mcmc.3, ffpk.mcmc.4))
dim(sims)
```

The **bayesplot** package can use to visualize output.


```{r}
names(ffpk.mcmc.1)
parms_name <- c("Fgutabs.1.", "kgutabs.1.", "kelim.1.", "Vdist.1.")
mcmc_trace(sims, pars = parms_name, facet_args = list(ncol = 1, strip.position = "left"))
```


```{r}
mcmc_dens_overlay(x = sims[1001:2000,,], pars = parms_name)
```


```{r}
mcmc_pairs(sims[1001:2000,,], pars = parms_name, off_diag_fun = "hex")
```


```{r}
monitor(sims[,,parms_name], digit=4) 
```


Use setpoints
```{r}
ffpk.setpts.out <- mcsim("models/ffpk.model", input = "inputs/ffpk.setpts.in")
tail(ffpk.setpts.out)
```

```{r}
t <- Theoph$Time[1:11]
obs<- Theoph$conc[1:11]
str.pt <- which(names(ffpk.setpts.out)=="Conc_1.1") 
end.pt <- which(names(ffpk.setpts.out)=="Conc_1.11") 
par(mfrow = c(1,1))
plot(t, obs, pch = "")
for(i in 501:1000){
  lines(t, ffpk.setpts.out[i,str.pt:end.pt], col="grey")
}
points(t, obs)
```

